Owlifier: Creating OWL-DL Ontologies from Simple Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}

### 2008-12-03 docs/pubs/ISEI6/main.tex

While not as expressive as OWL-DL, our approach can produce ontology
structures that are essential for improved data discovery and
integration \cite{madin07:_ontol_for_descr_and_synth}. Just as
important, because spreadsheet tools are frequently used by ecologists
to store and analyze data, \owlifier\ can provide ecologists with a
familiar and accessible user interface for ontology creation. This
approach also leverages the ability of spreadsheet tools to organize
and manipulate tabular data, e.g., allowing users to rapidly construct
class hierarchies from long lists of keywords.  In this way, an
ecologist can easily construct (or import) a set of terms, and then
incrementally organize these into class hierarchies, properties, and
constraints. In initial experiments with ecologists and evolutionary
biologists studying trait data, we found that \owlifier\ enabled them
to quickly and easily comprehend and construct useful ontologies.

The rest of this paper is organized as follows. In
\secref{sec:owlifier} we describe the basic syntax and semantics of
\owlifier. We define blocks that support a large subset of OWL-DL and
that also generally follow the ontology creation guidelines defined in
\cite{rector04:_owl_pizzas}. We also simplify certain aspects of
ontology creation using OWL-DL, e.g., by assuming classes are disjoint
by default (unless specified otherwise) and supporting sufficient
restrictions \cite{rector04:_owl_pizzas}. In \secref{sec:features} we
describe additional characteristics of \owlifier\ and discuss issues
with respect to classification and reasoning. In
\secref{sec:implementation} we briefly describe the \owlifier\
implementation, and conclude in \secref{sec:conclusion} with related
and future work. In general, the goal of \owlifier\ is not to support
all constructs in OWL-DL, but instead to provide a higher-level
ontology syntax (via spreadsheet blocks) that is easy for ecologists
to use and understand while also providing the necessary constructs
for developing typical ecological ontologies. By compiling \owlifier\
to OWL-DL, we also allow for experts to refine and extend the ontology
using more advanced ontology editing tools if necessary
(cf.~\figref{fig:owlifier}).

@@

\myblock{Transitive Blocks.} Transitive blocks are special cases of
relationship blocks where the object property is asserted to be
transitive. If a property $p$ is declared to be transitive, whenever
$p$ relates an individual $x$ to an individual $y$, and an
individual $y$ to an individual $z$, then $p$ is also assumed to
relate $x$ to $z$. Rows containing transitive blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{transitive} & $p$ & $c_1$
    & $c_2$ & \dots & $c_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $p$ is an object property and each $c$ is a class.  The
following block is a simple example of a transitive relationship.
\begin{tabbing}
  ~~\textsf{transitive} \textsf{hasPart} \textsf{Body}
  \textsf{Head} \textsf{Eye} \textsf{Retina}
\end{tabbing} 
This block states that every instance of the class Body has a Head as
a part, every instance of the class Head has an Eye as a part, and
every instance of the class Eye has a Retina as a part. Moreover,
because the \textsf{hasPart} property above is defined to be
transitive, it is possible to infer that every instance of Body also
has an Eye and a Retina as a part through the inherited relationship
restrictions $\textsf{Body} \sqsubseteq \exists \textsf{hasPart}
. \textsf{Head}$, $\textsf{Head} \sqsubseteq \exists \textsf{hasPart}
. \textsf{Eye}$, and $\textsf{Eye} \sqsubseteq \exists
\textsf{hasPart} . \textsf{Retina}$.

\myblock{Cardinality Blocks.} Cardinality blocks are also similar to
relationship blocks.  We consider three types of cardinality blocks
for defining minimum, maximum, and exact cardinality restrictions. A
minimum block states the smallest number of properties $p$ to distinct
individuals that an individual of a class may have.  Rows containing
minimum blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{min} & $p$ & $m$ & 
    $c_1$ & $c_2$ & $\dots$ & $c_n$
    \\ \hline 
  \end{tabular} \hfill $(n \ge 2)$
\end{itemize}
where $m$ is the minimum number of properties $p$ that instances of
class $c_i$ must have to instances of concept $c_{i+1}$, for $1 \le i
< n$.  For each class $c_i$, a minimum cardinality block induces the
description logic axiom $c_{i} \sqsubseteq \; (\le m) \; p.c_{i+1}$
stating that each instance of $c_i$ must be $p$-related to at least
$m$ unique instances of $c_{i+1}$. The following two blocks
demonstrate simple minimum cardinality constraints.
\begin{tabbing}
  ~~\textsf{min} \textsf{hasPart} 1 \textsf{Body} \textsf{Head} 
  \textsf{Nose} \\
  ~~\textsf{min} \textsf{hasPart} 2 \textsf{Head} \textsf{Eye}
\end{tabbing}
The first block states that instances of the class Body have at least
one Head as a part, which in turn have at least one Nose as a
part.\footnote{Cardinality restrictions ensuring participation to at
  least one property are typically not given through minimum
  cardinality blocks since they are also implied by relationship
  blocks.} The second block states that instances of the class Head
have at least two Eyes as parts.

A maximum block states the largest number of properties $p$ to
distinct individuals that an individual of a class may have.  Rows
containing maximum blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{max} & $p$ & $m$ & 
    $c_1$ & $c_2$ & $\dots$ & $c_n$
    \\ \hline
  \end{tabular} \hfill $(n \ge 2)$
\end{itemize}
where $m$ is the maximum number of properties $p$ that instances of
concept $c_i$ may have to instances of concept $c_{i+1}$, for $1 \le i
< n$.  For each class $c_i$, a maximum cardinality block induces the
description logic axiom $c_i \sqsubseteq \; (\ge m) \; p.c_{i+1}$
stating that each instance of $c_i$ may be $p$-related to at most $m$
unique instances of $c_{i+1}$. The following two blocks demonstrate
simple maximum cardinality constraints.
\begin{tabbing}
  ~~\textsf{max} \textsf{hasPart} 1 \textsf{Body} \textsf{Head} 
  \textsf{Nose} \\
  ~~\textsf{max} \textsf{hasPart} 2 \textsf{Head} \textsf{Eye}
\end{tabbing}
The first block states that instances of the class Body have at most
one Head as a part, which in turn has at most one Nose as a part. The
second block states that instances of the class Head have at most two
Eyes as parts.

An exact block ensures both a minimum and maximum number $m$ of
properties $p$ to distinct individuals that an individual of a class
must have. Rows containing exact blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{exact} & $p$ & $m$ & 
    $c_1$ & $c_2$ & $\dots$ & $c_n$
    \\ \hline
  \end{tabular} \hfill $(n \ge 2)$
\end{itemize}
where $m$ is the number of properties $p$ that instances of concept
$c_i$ must have to instances of concept $c_{i+1}$, for $1 \le i < n$.
For each class $c_i$, an exact block induces the description logic
axiom $c_i \sqsubseteq \; (= m) \; p.c_{i+1}$ stating that each
instance of $c_i$ must be $p$-related to $m$ unique instances of
$c_{i+1}$.

@@

\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym
blocks in that they state equivalences between classes. A sufficient
block serves as a modifier to entity blocks, relationship blocks
(including transitive blocks), and cardinality blocks, modifying the
associated description logic axioms by using equivalence ($\equiv$) in
place of subclass ($\sqsubseteq$). For example the block
\begin{tabbing}
  ~~\textsf{sufficient} \textsf{relationship} \textsf{hasPart} \textsf{Mammal} \textsf{Hair} 
\end{tabbing}
induces the description logic axiom $\textsf{Mammal} \equiv \exists
\textsf{hasPart}.\textsf{Hair}$ stating that \emph{any} individual
that has Hair as a part is a Mammal. Additionally, we allow
relationship blocks within a sufficient block to be extended with
\textsf{not} to state that absence of the property is a defining
characteristic of the class. For example the following blocks
\begin{tabbing}
  ~~\textsf{sufficient} \textsf{entity} \textsf{Mammal} \textsf{Eutheria} \\
  ~~\textsf{sufficient} \textsf{relationship not} \textsf{hasPart} \textsf{Eutheria} \textsf{EpipubicBone} 
\end{tabbing}
induces the axiom $\textsf{Eutheria} \equiv \textsf{Mammal} \sqcap
\neg \exists \textsf{hasPart} . \textsf{EpipubicBone}$ stating that
any Mammal that does not have an Epipubic Bone as a part is a
Eutheria. Note also that multiple sufficient blocks for a particular
class result in a single axiom in which each constraint is combined
via conjunction ($\sqcap$).

\myblock{Comment Blocks.} Comment blocks provide a mechanisms to add
plain-text comments to \owlifier\ tables. A description block attaches
a plain-text comment to classes and properties. Rows containing
description blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline \textsf{description} & $c$ or $p$ & $s$
    \\ \hline
  \end{tabular}
\end{itemize}
where the string $s$ is associated as a comment to the class $c$ or
property $p$. A note block allows attaches comments to the current
ontology. Rows containing note blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline \textsf{note} & $s$
\\ \hline
  \end{tabular}
\end{itemize}
where $s$ is a comment string.

@@

In this section, we briefly describe some of the additional features
of \owlifier, specifically focusing on the use of disjoint classes,
\owlifier\ reasoning, and additional block syntax.

@@

OWL is based on the open world assumption, which can lead to a number
of ontology development ``pitfalls'' for those new to the language
\cite{smith04:_owl_web_ontol_languag_guide,rector04:_owl_pizzas}. One
example is in the creation of disjoint classes.  In particular, unless
explicitly stated, distinct classes within an OWL-DL ontology are
never assumed to be disjoint. However, in many ontologies a large
number of classes are typically defined as being disjoint (e.g.,
sibling classes), and stating these disjoint constraints is often
time-consuming since each pair of classes must be given an explicit
disjoint assertion. Editors such as Protege
\cite{knublauch04:_editin_descr_logic_ontol_with} provide shortcuts
via the user interface to create specific sets of disjoint assertions,
e.g., by allowing a user to define all children of a particular class
as disjoint. In general, however, many users expect such classes to be
disjoint by default \cite{rector04:_owl_pizzas} and this expectation
often leads to modeling errors.

@@

As described in \cite{rector04:_owl_pizzas} additional reasoning can
occur within OWL-DL ontologies when domain and range axioms are
provided (as well as, e.g., property closure axioms).  We explicitly
do not consider these constraints in the current version of \owlifier\
because they also often result in modeling errors for inexperienced
OWL-DL users \cite{rector04:_owl_pizzas}. Instead, we adopt the more
traditional description-logic approaches that do not have explicit
domain and range axioms.  Although not currently supported in
\owlifier, domain and range constraints could be inferred from given
relationship blocks as well as property closure axioms.

@@

To help simplify the creation of class hierarchies and property
sequences (including transitive, cardinality, and sufficient blocks),
we allow for ``sparse'' blocks that inherit missing information from
their closest proceeding block. For instance the following entity
blocks
\begin{tabbing}
  ~~\textsf{entity} \= \textsf{EcologicalHabitat} \= \textsf{Estuary} \= \kill
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \> 
  \textsf{Bay}\\
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \>
  \textsf{Fjord} \\
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Marsh} \>
  \textsf{TidalMarsh} \\ 
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Marsh} \>
  \textsf{SaltMarsh}
\end{tabbing}
can be equivalently represented in \owlifier\ using the following
sparse blocks.
\begin{tabbing}
  ~~\textsf{entity} \= \textsf{EcologicalHabitat} \= \textsf{Estuary} \= \kill
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \> 
  \textsf{Bay}\\
  \> \> \> \textsf{Fjord} \\
  \> \> \textsf{Marsh} \> \textsf{TidalMarsh} \\ 
  \> \> \> \textsf{SaltMarsh}
\end{tabbing}
In general, the use of sparse blocks simplifies ontology creation by
not requiring users to enter every (redundant, leading) field to be
explicitly given, which in turn can simplify the overall layout of the
ontology within a spreadsheet.  Additionally, \owlifier\ does not place
constraints on the order of blocks within a spreadsheet. Classes also
do not need to explicitly be defined within an entity block, e.g.,
classes without corresponding entity blocks can be introduced simply
through synonym and relationship blocks. This typically occurs when a
particular class does not participate as a subclass or superclass of
another class in the current ontology. Overall, one of our goals is
for \owlifier\ to allow minimal information to be entered by limiting
redundancy as well as by supporting a range of inferences, while at
the same time reducing the number of errors commonly made in ontology
development by non-experts.

\section{The \Owlifier Application}
\label{sec:implementation}

\figref{fig:owlifier} shows the general architecture of the \owlifier\
application. A scientist or researcher first creates a spreadsheet
containing a set of ontology definitions using \owlifier\ conventions.
The spreadsheet is then exported as a plain-text file containing the
\owlifier\ table, e.g., using a CSV or tab-delimited format. The
\owlifier\ table is then sent to the \owlifier\ application, which
performs a number of steps that include: (i) parsing the file; (ii)
generating the corresponding OWL-DL representation (e.g., in the
current implementation, the Jena API is used \cite{carroll04:_jena});
(iii) validating the ontology and ensuring it is consistent via an
OWL-DL reasoner (e.g., the current implementation uses Pellet
\cite{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL
file. The resulting OWL file can then be refined and extended via
existing OWL tools (e.g., Protege or SWOOP). It is possible that the
refined ontology is converted back to a corresponding \owlifier\
representation (shown using dashed lines in
\figref{fig:owlifier}). While simple to perform, this last conversion
is not information preserving since not all OWL-DL constructs are
represented through \owlifier\ blocks.

The current implementation of the \owlifier\ application is written in
Java and supports a subset of the blocks defined in
\secref{sec:owlifier}. In particular, we are currently extending
\owlifier\ to support cardinality and sufficient blocks as well as to
perform the above conversion from OWL-DL files to corresponding
\owlifier\ blocks. As noted above, the current implementation,
although it does not support all of the blocks defined here, is being
used within a project focused on supporting trait data and many of the
extensions outlined here are based on these experiences. In addition,
\owlifier\ has been used to construct term hierarchies from large sets
of keywords (harvested from existing metadata documents) as well as to
define articulations into the observation ontology defined in
\cite{madin07:_ontol_for_descr_and_synth}.

@@

@@

In this paper, we have presented a new approach for developing
ontologies to address barriers in ontology development and adoption by
ecologists. Our approach allows scientists to use familiar spreadsheet
software (frequently used, e.g., by ecologists for storing and
analyzing data) to create ontologies by filling in a set of templates,
or blocks, that generate OWL-DL class hierarchies, properties, and
constraints via the \owlifier\ application.  This approach can provide
a more intuitive and accessible ontology editing environment for
ecologists, especially compared to existing OWL-based tools that
require considerable expertise in the underlying logical formalisms.

Protege provides a variety of ontology editing plug-ins, including a
simple interface for text-based editing of class hierarchies. In
\cite{kola07:_impor_spread_data_into_proteg}, an approach is described
for importing spreadsheet-based ontology descriptions into
Protege. However, this approach aims at supporting \emph{ad-hoc}
spreadsheet structures and then provides an intermediate step and
interface for mapping these structures into ontology axioms. This
approach is similar to others (e.g.,
\cite{han08:_rdf12,bizer03:_d2r_map,an06:_discov_seman_of_relat_tables_throug_mappin})
for defining mappings from relational data to RDF and OWL ontology
class and instance data. To the best of our knowledge, however, our
approach is the first to consider a detailed set of templates that can
support a large subset of existing OWL constructs. In addition, we
define a number of shortcuts for creating \owlifier\ tables, including
sparse blocks, as well as default semantics (e.g., disjoint classes,
which is a form of global taxonomic constraint defined in
\cite{thau07:_reason_about_taxon_in_first_order_logi}) that further
simplify ontology creation for end users.

>>>>

While not as expressive as OWL-DL, our approach can produce ontology
structures that are essential for improved data discovery and
integration \cite{madin07:_ontol_for_descr_and_synth}. Just as
important, because spreadsheet tools are frequently used by ecologists
to store and analyze data, \owlifier\ can provide ecologists with a
familiar and accessible user interface for ontology creation. This
approach also leverages the ability of spreadsheet tools to organize
and manipulate tabular data, e.g., allowing users to rapidly construct
class hierarchies from long lists of keywords.  In this way, an
ecologist can easily construct (or import) a set of terms, and then
incrementally organize these into class hierarchies, properties, and
constraints. In initial experiments with ecologists and evolutionary
biologists studying trait data, we found that \owlifier\ enabled them
to quickly and easily comprehend and construct relatively complex and
meaningful ontologies.

The rest of this paper is organized as follows. In
\secref{sec:owlifier} we describe the basic syntax and semantics of
\owlifier. We define blocks that support a large subset of OWL-DL and
that also generally follow the ontology creation guidelines defined in
\cite{rector04:_owl_pizzas}. We also simplify certain aspects of
ontology creation using OWL-DL, e.g., by assuming classes are disjoint
by default (unless specified otherwise)
\cite{rector04:_owl_pizzas}. In \secref{sec:features} we describe
additional characteristics of \owlifier\ and discuss issues with
respect to classification and reasoning. In
\secref{sec:implementation} we briefly describe the \owlifier\
implementation, and conclude in \secref{sec:conclusion} with related
and future work. In general, the goal of \owlifier\ is not to support
all constructs in OWL-DL, but instead to provide a higher-level
ontology syntax (via spreadsheet blocks) that is easy for ecologists
to use and understand while also providing the necessary constructs
for developing typical ecological ontologies. By compiling \owlifier\
to OWL-DL, we also allow for experts to refine and extend the ontology
using more advanced ontology editing tools if necessary
(cf.~\figref{fig:owlifier}).

@@

\myblock{Transitive Blocks.} Transitive blocks are special cases of
relationship blocks where the object property is asserted to be
transitive. That is, if a property $p$ is declared to be transitive,
and $p$ relates an individual $x$ to an individual $y$ and $y$ to an
individual $z$, then $p$ is assumed to also relate $x$ to $z$. Rows
containing transitive blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{transitive} & $p$ & $c_1$
    & $c_2$ & \dots & $c_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $p$ is an object property and each $c$ is a class.  The
following block is a simple example of a transitive relationship
definition.
\begin{tabbing}
  ~~\textsf{transitive} \textsf{hasPart} \textsf{Body}
  \textsf{Head} \textsf{Eye} \textsf{Retina}
\end{tabbing} 
This block states that every instance of the class Body has a Head as
a part, every instance of the class Head has an Eye as a part, and
every instance of the class Eye has a Retina as a part. Moreover,
because the \textsf{hasPart} property above is defined to be
transitive, it is possible to infer that every instance of Body also
has an Eye and a Retina as a part through the inherited relationship
restrictions $\textsf{Body} \sqsubseteq \exists \textsf{hasPart}
. \textsf{Head}$, $\textsf{Head} \sqsubseteq \exists \textsf{hasPart}
. \textsf{Eye}$, and $\textsf{Eye} \sqsubseteq \exists
\textsf{hasPart} . \textsf{Retina}$.

\myblock{Cardinality Blocks.} Cardinality blocks are also similar to
relationship blocks.  We consider three types of cardinality blocks
for defining minimum, maximum, and exact cardinality restrictions.
Rows containing minimum blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{min} & $p$ & $m$ & 
    $c_1$ & $c_2$ & $\dots$ & $c_n$
    \\ \hline 
  \end{tabular} \hfill $(n \ge 2)$
\end{itemize}
where $m$ is the minimum number of properties $p$ that instances of
class $c_i$ must have to instances of concept $c_{i+1}$, for $1 \le i
< n$.  For each class $c_i$, a minimum cardinality block induces the
description logic axiom $c_{i} \sqsubseteq \; (\le m) \; p.c_{i+1}$
stating that each instance of $c_i$ must be $p$-related to at least
$m$ unique instances of $c_{i+1}$. The following two blocks
demonstrate simple minimum cardinality constraints.
\begin{tabbing}
  ~~\textsf{min} \textsf{hasPart} 1 \textsf{Body} \textsf{Head} 
  \textsf{Nose} \\
  ~~\textsf{min} \textsf{hasPart} 2 \textsf{Head} \textsf{Eye}
\end{tabbing}
The first block states that instances of the class Body have at least
one Head as a part, which in turn have at least one Nose as a
part.\footnote{Cardinality restrictions ensuring participation to at
  least one property are typically not given through minimum
  cardinality blocks since they are also implied by relationship
  blocks.} The second block states that instances of the class Head
have at least two Eyes as parts.

Rows containing maximum blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{max} & $p$ & $m$ & 
    $c_1$ & $c_2$ & $\dots$ & $c_n$
    \\ \hline
  \end{tabular} \hfill $(n \ge 2)$
\end{itemize}
where $m$ is the maximum number of properties $p$ that instances of
concept $c_i$ may have to instances of concept $c_{i+1}$, for $1 \le i
< n$.  For each class $c_i$, a maximum cardinality block induces the
description logic axiom $c_i \sqsubseteq \; (\ge m) \; p.c_{i+1}$
stating that each instance of $c_i$ may be $p$-related to at most $m$
unique instances of $c_{i+1}$. The following two blocks demonstrate
simple maximum cardinality constraints.
\begin{tabbing}
  ~~\textsf{max} \textsf{hasPart} 1 \textsf{Body} \textsf{Head} 
  \textsf{Nose} \\
  ~~\textsf{max} \textsf{hasPart} 2 \textsf{Head} \textsf{Eye}
\end{tabbing}
The first block states that instances of the class Body have at most
one Head as a part, which in turn has at most one Nose as a part. The
second block states that instances of the class Head have at most two
Eyes as parts.

Rows containing exact blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{exact} & $p$ & $m$ & 
    $c_1$ & $c_2$ & $\dots$ & $c_n$
    \\ \hline
  \end{tabular} \hfill $(n \ge 2)$
\end{itemize}
where $m$ is the number of properties $p$ that instances of concept
$c_i$ must have to instances of concept $c_{i+1}$, for $1 \le i < n$.
For each class $c_i$, an exact block induces the description logic
axiom $c_i \sqsubseteq \; (= m) \; p.c_{i+1}$ stating that each
instance of $c_i$ must be $p$-related to $m$ unique instances of
$c_{i+1}$.

@@

\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym
blocks in that they state equivalences between classes. A sufficient
block serves as a modifier to entity blocks, relationship blocks
(including transitive blocks), and cardinality blocks, modifying the
associated description logic axioms by using equivalence ($\equiv$) in
place of subclass ($\sqsubseteq$). For example the block
\begin{tabbing}
  ~~\textsf{sufficient} \textsf{relationship} \textsf{hasPart} \textsf{Mammal} \textsf{Hair} 
\end{tabbing}
induces the description logic axiom $\textsf{Mammal} \equiv \exists
\textsf{hasPart}.\textsf{Hair}$ stating that \emph{any} individual
that has Hair as a part is a Mammal. Additionally, we allow
relationship blocks within a sufficient block to be extended with
\textsf{not} to state that the absence of the property is a defining
characteristic of the class. For example the following blocks
\begin{tabbing}
  ~~\textsf{sufficient} \textsf{entity} \textsf{Mammal} \textsf{Eutheria} \\
  ~~\textsf{sufficient} \textsf{relationship not} \textsf{hasPart} \textsf{Eutheria} \textsf{EpipubicBone} 
\end{tabbing}
induces the axiom $\textsf{Eutheria} \equiv \textsf{Mammal} \sqcap
\neg \exists \textsf{hasPart} . \textsf{EpipubicBone}$ stating that
any Mammal that does not have an Epipubic Bone as a part is a
Eutheria. Note also that multiple sufficient blocks for a particular
class result in a single axiom in which each constraint (i.e., class
constructor) is combined via intersection ($\sqcap$).

\myblock{Comment Blocks.} Comment blocks provide a mechanism to add
plain-text comments to \owlifier\ tables. A description block attaches
a plain-text comment to classes and properties. Rows containing
description blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline \textsf{description} & $c$ or $p$ & $s$
    \\ \hline
  \end{tabular}
\end{itemize}
where the string $s$ is associated as a comment to the class $c$ or
property $p$. A note block attaches comments to the current
ontology. Rows containing note blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline \textsf{note} & $s$
\\ \hline
  \end{tabular}
\end{itemize}
where $s$ is a comment string.

@@

Here we briefly describe some of the additional features of \owlifier,
specifically focusing on the use of disjoint classes, \owlifier\
reasoning, and additional block syntax.

@@

OWL is based on the open world assumption, which can lead to a number
of ontology development ``pitfalls'' for those new to the language
\cite{smith04:_owl_web_ontol_languag_guide,rector04:_owl_pizzas}. One
example is in the creation of disjoint classes.  In particular, unless
explicitly asserted, distinct classes within an OWL-DL ontology are
not assumed to be disjoint. However, in many ontologies a large number
of classes are typically defined as being disjoint (e.g., sibling
classes), and stating these disjoint constraints is often
time consuming since each pair of classes must be given an explicit
disjoint assertion. Editors such as Protege
\cite{knublauch04:_editin_descr_logic_ontol_with} provide shortcuts
via the user interface to create specific sets of disjoint assertions,
e.g., by allowing a user to define all children of a particular class
as disjoint. In general, however, many users expect such classes to be
disjoint by default \cite{rector04:_owl_pizzas} and this expectation
often leads to modeling errors.

@@

As described in \cite{rector04:_owl_pizzas} additional reasoning can
occur within OWL-DL ontologies when domain and range axioms are
provided (as well as, e.g., property closure axioms).  We explicitly
do not consider these constraints in the current version of \owlifier\
because they also often result in modeling errors for inexperienced
OWL-DL users \cite{rector04:_owl_pizzas}. Instead, we adopt the
approach of more traditional description logics, which do not have
explicit domain and range axioms.  Although not currently supported in
\owlifier, domain and range constraints as well as property closure
axioms can be inferred from given relationship blocks.

@@

To help simplify the creation of class hierarchies and property
sequences (including transitive, cardinality, and sufficient blocks),
we allow for ``sparse'' blocks that inherit missing information from
their closest proceeding block. For instance the following entity
blocks
\begin{tabbing}
  ~~\textsf{entity} \= \textsf{EcologicalHabitat} \= \textsf{Estuary} \= \kill
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \> 
  \textsf{Bay}\\
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \>
  \textsf{Fjord} \\
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Marsh} \>
  \textsf{TidalMarsh} \\ 
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Marsh} \>
  \textsf{SaltMarsh}
\end{tabbing}
can be equivalently represented in \owlifier\ using the following
sparse blocks.
\begin{tabbing}
  ~~\textsf{entity} \= \textsf{EcologicalHabitat} \= \textsf{Estuary} \= \kill
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \> 
  \textsf{Bay}\\
  \> \> \> \textsf{Fjord} \\
  \> \> \textsf{Marsh} \> \textsf{TidalMarsh} \\ 
  \> \> \> \textsf{SaltMarsh}
\end{tabbing}
In general, the use of sparse blocks simplifies ontology creation by
not requiring users to enter every redundant field explicitly, which
in turn can simplify the overall layout of the ontology within a
spreadsheet.  Additionally, \owlifier\ does not place constraints on
the order of blocks within a spreadsheet. Classes also do not need to
be explicitly defined within an entity block, e.g., classes without
corresponding entity blocks can be introduced simply through synonym
and relationship blocks. This typically occurs when a particular class
does not participate as a subclass or superclass of another class in
the current ontology. Taken together, these appraoches allow users to
enter minimal information by limiting redundancy and by supporting a
range of inferences, while at the same time reducing the causes of
many errors commonly made in ontology development by non-experts.

\section{The \Owlifier\ Implementation}
\label{sec:implementation}

\figref{fig:owlifier} shows the general architecture of the \owlifier\
application. A scientist or researcher first creates a spreadsheet
containing a set of ontology definitions using \owlifier\ conventions.
The spreadsheet is then exported as a plain-text file containing the
\owlifier\ table, e.g., using a CSV or tab-delimited format. The
\owlifier\ table is then sent to the \owlifier\ application, which
performs a number of steps that include: (i) parsing the file; (ii)
generating the corresponding OWL-DL representation (e.g., in the
current implementation, the Jena API is used \cite{carroll04:_jena});
(iii) validating the ontology and ensuring it is consistent via an
OWL-DL reasoner (e.g., the current implementation uses Pellet
\cite{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL
file. The resulting OWL file can then be refined and extended via
existing OWL tools (e.g., Protege or SWOOP). It is possible that the
refined ontology is converted back to a corresponding \owlifier\
representation (shown using dashed lines in
\figref{fig:owlifier}). While straightforward to perform, this last
conversion is not information preserving since not all OWL-DL
constructs are represented through \owlifier\ blocks.

@@

The current implementation of the \owlifier\ application is written in
Java and supports a subset of the blocks defined in
\secref{sec:owlifier}. In particular, we are currently extending
\owlifier\ to support cardinality and sufficient blocks as well as to
perform the above conversion from OWL-DL files to corresponding
\owlifier\ blocks. As noted above, the current implementation,
although it does not support all of the blocks defined here, is being
used within a project focused on supporting trait data and many of the
extensions outlined here are based on these experiences. In addition,
\owlifier\ has been used to construct term hierarchies from large sets
of keywords (harvested from existing metadata documents) as well as to
define articulations between existing ontologies and the observation
ontology defined in \cite{madin07:_ontol_for_descr_and_synth}.

@@

This paper presents a new approach for developing ontologies to
address barriers in ontology development and adoption by
ecologists. Our approach allows scientists to use familiar spreadsheet
software (e.g., frequently used by ecologists for storing and
analyzing data) to create ontologies by filling in a set of templates,
or blocks, that generate OWL-DL class hierarchies, properties, and
constraints via the \owlifier\ application.  This approach can provide
a more intuitive and accessible ontology editing environment for
ecologists, especially compared to existing OWL-based tools that
require considerable expertise in the underlying logic formalisms.

Protege provides a variety of ontology editing plug-ins, including a
simple interface for text-based editing of class hierarchies. In
\cite{kola07:_impor_spread_data_into_proteg}, an approach is described
for importing spreadsheet-based ontology descriptions into
Protege. However, this approach aims at supporting \emph{ad-hoc}
spreadsheet structures by providing an intermediate interface for
mapping these structures into ontology axioms. This approach is
similar to others (e.g.,
\cite{han08:_rdf12,bizer03:_d2r_map,an06:_discov_seman_of_relat_tables_throug_mappin})
for defining mappings from relational data to RDF and OWL ontology
class and instance data. To the best of our knowledge, however, our
approach is the first to consider a detailed set of templates that can
support a large subset of existing OWL constructs. In addition, we
define a number of shortcuts for creating \owlifier\ tables, including
sparse blocks and default semantics (e.g., disjoint classes, which is
a form of global taxonomic constraint defined in
\cite{thau07:_reason_about_taxon_in_first_order_logi}) that further
simplify ontology creation for end users.
